Mallows, Black-box combinatorial optimization, limited budget, ???

### 2020-10-23 paper/main.tex
more edits

In this work, we propose and analyze UMM, an EDA specifically designed for
black-box combinatorial optimization under a limited budget. The proposed UMM 
is a population based algorithm. There exists a wide body of research on population based probabilistic algorithms, where most prominent examples are Estimation of Distribution Algorithms (EDAs). Our approach drastically differs from EDAs since EDAs assume that function evaluations are cheap and they can do many of them. On the contrary, UMM minimizes the number of function evaluations. Moreover, UMM does not rely on an evolving sequence of samples (called generations in evolutionary approaches). Instead, UMM considers a single sample of permutations during the whole execution that is increased iteratively, with one permutation each iteration.

@@

Let us assume a black-box ``fitness'' function that must be minimized over a
space of permutations $f\colon S_n \to \mathbb{R}$. Being black-box means that
we can evaluate any candidate permutation $\sigma \in S_n$ to obtain
$f(\sigma)$, however, we do not know anything else about the form of
$f$. Moreover, the evaluation of $f$ is expensive in computation time or
resources and, thus, limits the budget of candidate permutations that we can
evaluate to a relatively low number. Here, we will study budgets lower than
400. Such expensive black-box combinatorial problems arise, for example, in
manufacturing~\citep{FerAlvDiaIglEna2014ants}, where the fitness function
may involve expensive simulations for which no closed-form mathematical
description is available.

@@

@@

A purely explotation approach\ekhine{esto se dice asi? uno greedy igual esta mejor dicho} would systematically generate $\sigma_0$ in the sampling stage.\MANUEL{I'm not sure what ``systematically generate $\sigma_0$'' means.} However, this setting\MANUEL{which setting? $\theta$?}  controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{arza2019approaching}.\MANUEL{there is something wrong with this reference\ekhine{}{qué le pasa?} Don't you see a complete mess of the authors list? It looks like it has 20 authors.}

@@

LOP: Ekhine

Quizás luego: PFSP y QAP

@@

Complete this with details of UMM: MANUEL

@@

@@

In the case of the QAP and PFSP, we consider the same instances as
\citet{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, i.e.,
\texttt{nug12},\texttt{nug30},\texttt{tho30} and \texttt{kra32} for the QAP,
with $n$ equal to the number in the instance name, and \texttt{reC05},
\texttt{reC13}, \texttt{reC19}, \texttt{reC31} for the PFSP, with
$n \in \{20, 20, 30, 50\}$, respectively.

@@

 \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean of the difference between the fitness of UMM minus the one of CEGO. Mean runtime is measured in hours. If a run went over 72 hours, of  CEGO was not able to reach 400 function evaluations (FEs) in 72 hours, it was terminated earlier. This was the case for all LOP instances shown.\label{tab:results}}

@@

\begin{tabular}{*{6}{r}rl*{3}{r}}

@@

            &                  & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI for the\\mean difference}}} & \bf \multirow[b]{2.45}{*}{\shortstack{Mean FE\\CEGO}} & \mcolc{2}{Mean runtime} \\\cmidrule{3-6}\cmidrule(lr){10-11}

@@

    LOP     & N-t1d100.02      & 79686.5                 & (367.9)    & 76827.7   & (891.0)                & $[$2194.5,    & 3523.1$]$     & 110.7& 72.9 & 3.2  \\
    LOP     & N-t1d150.01      & 187605.5                & (1906.9)   & 179508.3  & (1647.9)               & $[$6420.3,    & 9774.1$]$     & 73.9 & 72.7 & 6.5  \\
    LOP     & N-t1d150.02      & 184320.9                & (973.5)    & 177592.8  & (1196.2)               & $[$5700.5,    & 7755.7$]$     & 74.1 & 73.2 & 6.5  \\
    LOP     & N-t1d200.01      & 335391.1                & (2289.0)   & 323513.6  & (1371.2)               & $[$10076.1,   & 13678.9$]$    & 55.6 & 73.2 & 10.7 \\
    LOP     & N-t1d200.02      & 333149.5                & (2497.2)   & 319675.8  & (2242.6)               & $[$11242.0,   & 15705.4$]$    & 56.5 & 73.7 & 10.6 \\
    LOP     & N-t2d150.01      & 24682.7                 & (976.6)    & 15297.7   & (563.5)                & $[$8622.2,    & 10147.8$]$    & 74.6 & 72.9 & 7.0  \\
    LOP     & N-t2d150.02      & 24250.8                 & (1354.9)   & 14724.9   & (667.5)                & $[$8495.1,    & 10556.7$]$    & 74.8 & 72.8 & 7.1  \\
    LOP     & N-t2d200.01      & 57098.6                 & (4452.8)   & 33465.2   & (2001.0)               & $[$20284.5,   & 26982.3$]$    & 56.3 & 73.3 & 11.2 \\
    LOP     & N-t2d200.02      & 54895.6                 & (4589.7)   & 35901.8   & (2077.5)               & $[$15539.1,   & 22448.5$]$    & 55.6 & 73.5 & 11.2 \\
    LOP     & N-p50-01         & 21431.2                 & (299.5)    & 21027.6   & (711.8)                & $[$-128.1,    & 935.3$]$      & 219.9& 72.3 & 0.9  \\
    LOP     & N-p50-02         & 22021.1                 & (275.6)    & 21613.6   & (596.4)                & $[$-42.5,     & 857.5$]$      & 217.3& 72.3 & 0.9  \\
    LOP     & N-atp111         & 693.1                   & (10.3)     & 594.0     & (15.4)                 & $[$86.7,      & 111.5$]$      & 98.7 & 72.5 & 4.1  \\
    LOP     & N-atp134         & 880.5                   & (18.6)     & 756.3     & (21.3)                 & $[$105.4,     & 143.0$]$      & 83.0 & 72.8 & 5.7  \\
    LOP     & N-be75eec\_150   & 1666193.1               & (72310.8)  & 1415208.5 & (58593.5)              & $[$188960.6,  & 313008.6$]$   & 74.0 & 72.9 & 6.9  \\
    LOP     & N-be75eec\_250   & 5214413.0               & (166701.4) & 4431271.0 & (109756.7)             & $[$649040.8,  & 917243.2$]$   & 44.9 & 73.7 & 15.5 \\
    LOP     & N-be75np\_150    & 3742444.0               & (88500.2)  & 3204878.3 & (80704.1)              & $[$457944.2,  & 617187.2$]$   & 74.2 & 72.9 & 6.9  \\
    LOP     & N-be75np\_250    & 10887427.8              & (159642.5) & 9544754.9 & (260523.9)             & $[$1136635.8, & 1548710.0$]$  & 45.3 & 74.0 & 15.5 \\
   PFSP     & rec05.txt        & 1317.9                  & (25.1)     & 1330.7    & (17.9)                 & $[$-33.4,     & 7.8$]$        & 400.0& 38.9 & 0.1  \\
   PFSP     & rec13.txt        & 2128.2                  & (25.0)     & 2158.7    & (29.5)                 & $[$-56.2,     & -4.8$]$       & 400.0& 38.9 & 0.1  \\
   PFSP     & rec19.txt        & 2398.6                  & (19.8)     & 2405.1    & (12.6)                 & $[$-22.3,     & 9.3$]$        & 400.0& 82.1 & 0.3  \\
    QAP     & kra32.dat        & 115907.0                & (4663.4)   & 116320.0  & (4673.5)               & $[$-4799.3,   & 3973.3$]$     & 400.0& 95.8 & 0.3  \\
    QAP     & nug12.dat        & 624.2                   & (26.3)     & 660.4     & (17.3)                 & $[$-57.3,     & -15.1$]$      & 400.0& 14.8 & 0.0  \\
    QAP     & nug30.dat        & 7486.8                  & (40.4)     & 7474.0    & (83.9)                 & $[$-50.8,     & 76.4$]$       & 400.0& 84.3 & 0.3  \\
    QAP     & tho30.dat        & 192416.0                & (2584.2)   & 191482.0  & (4419.7)               & $[$-2527.0,   & 4395.0$]$     & 400.0& 85.1 & 0.3  \\

>>>>

In this work, we propose and analyze UMM, an EDA specifically designed for
black-box combinatorial optimization under a limited budget. The proposed UMM 
is a population based algorithm. There exists a wide body of research on population based probabilistic algorithms, where most prominent examples are Estimation of Distribution Algorithms (EDAs)~\citep{???}. Our approach drastically differs from EDAs since EDAs assume that function evaluations are cheap and they can do many of them. On the contrary, UMM minimizes the number of function evaluations. Moreover, UMM does not rely on an evolving sequence of samples (called generations in evolutionary approaches). Instead, UMM considers a single sample of permutations during the whole execution that is increased iteratively, with one permutation each iteration.

@@

@@

Let us assume a black-box ``fitness'' function that must be minimized over a
space of permutations $f\colon S_n \to \mathbb{R}$. Being black-box means that
we can evaluate any candidate permutation $\sigma \in S_n$ to obtain
$f(\sigma)$, however, we do not know anything else about the form of
$f$. Moreover, the evaluation of $f$ is expensive in computation time or
resources and, thus, we can only evaluate a limited budget of candidate
permutations. Here, we will study budgets lower than 400. Such expensive
black-box combinatorial problems arises in diverse contexts, for example,
protein folding~\citep{RomKraArn2012protein} and industrial
production~\citep{FerAlvDiaIglEna2014ants}, where the fitness function may
involve expensive simulations for which no closed-form mathematical description
is available.

@@

A purely explotation approach\ekhine{esto se dice asi? uno greedy igual esta mejor dicho} would systematically generate $\sigma_0$ in the sampling stage.\MANUEL{I'm not sure what ``systematically generate $\sigma_0$'' means in this context.} However, this setting\MANUEL{which setting? $\theta$?}  controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{arza2019approaching}.\MANUEL{there is something wrong with this reference\ekhine{}{qué le pasa?} Don't you see a complete mess of the authors list? It looks like it has 20 authors.}

@@

@@

@@

\subsubsection{Benchmark problems.} Experiments with real-world expensive black-box problems would be computationally infeasible. Instead, we consider classical combinatorial optimization problems as black-box optimization problems.

The \textbf{LOP} \ldots

The \emph{Quadratic Assignment Problem (QAP)} is defined by:
\begin{equation}\label{eq:qap}
  \min_{\sigma \in S_n} f(\sigma) = \sum_{i=1}^n \sum_{j=1}^n a_{i,j} b_{\sigma(i), \sigma(j)}
\end{equation}
%
where $[a_{i,j}]$ and $[b_{i,j}]$ are two matrices of size $n \times n$.

The \emph{Permutation Flowshop Scheduling Problem (PFSP)} is defined by a matrix $[p_{ij}]$ of size $n \times M$ that gives the processing time of a job $i \in [n]$ on a machine $j \in [M]$. All jobs must be processed by all machines in the same order. Given a permutation of the jobs $\sigma$, $C_{i,j}$ denotes the completion time of job at position $\sigma(i)$ on machine $j$, and $C_{\max} = C_{n,M}$  is the completion time of the last job on the last machine, i.e., the makespan. The objective of the PFSP is to find:
\begin{equation}\label{eq:qap}
  \begin{split}
    \min_{\sigma \in S_n}\quad& f(\sigma) = C_{n,M}\\
    \text{s.t.}\quad & C_{1,j} = 0 \quad j\in [M], \qquad C_{i,1} = 0\quad i \in [n]\\
    &C_{i,j} = p_{\sigma(i),j} + \max\{C_{i-1, j}, C_{i, j-1}\}  \ \  i \in \{2,\dotsc,n\},\ j \in\{2,\dotsc, M\}\\
\end{split}
\end{equation}

@@

In the case of the QAP and PFSP, we consider the same instances as
\citet{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, i.e.,
\texttt{nug12},\texttt{nug30},\texttt{tho30} and \texttt{kra32} for the QAP,
with $n$ (the permutation length) equal to the number in the instance name, and \texttt{reC05},
\texttt{reC13}, \texttt{reC19}, \texttt{reC31} for the PFSP, with
$n \in \{20, 20, 30, 50\}$, respectively.

@@

 \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean of the difference between the fitness of UMM minus the one of CEGO. Mean runtime is measured in hours. If a run went over 72 hours, of  CEGO was not able to reach 400 function evaluations (FEs) in 72 hours, it was terminated earlier. This was the case for all LOP instances shown.\manuel{}{Don't we have results for rec31?}\label{tab:results}}

@@

\begin{tabular}{r@{\hskip -2ex}*{5}{r}rl*{3}{r}}

@@

            &                  & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI for the\\mean difference}}} & \bf \multirow[b]{2.45}{*}{\shortstack{Mean FE\\CEGO}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){10-11}

@@

     & N-t1d100.02      & 79686.5                 & (367.9)    & 76827.7   & (891.0)                & $[$2194.5,    & 3523.1$]$     & 110.7& 72.9 & 3.2  \\
     & N-t1d150.01      & 187605.5                & (1906.9)   & 179508.3  & (1647.9)               & $[$6420.3,    & 9774.1$]$     & 73.9 & 72.7 & 6.5  \\
     & N-t1d150.02      & 184320.9                & (973.5)    & 177592.8  & (1196.2)               & $[$5700.5,    & 7755.7$]$     & 74.1 & 73.2 & 6.5  \\
     & N-t1d200.01      & 335391.1                & (2289.0)   & 323513.6  & (1371.2)               & $[$10076.1,   & 13678.9$]$    & 55.6 & 73.2 & 10.7 \\
     & N-t1d200.02      & 333149.5                & (2497.2)   & 319675.8  & (2242.6)               & $[$11242.0,   & 15705.4$]$    & 56.5 & 73.7 & 10.6 \\
     & N-t2d150.01      & 24682.7                 & (976.6)    & 15297.7   & (563.5)                & $[$8622.2,    & 10147.8$]$    & 74.6 & 72.9 & 7.0  \\
     & N-t2d150.02      & 24250.8                 & (1354.9)   & 14724.9   & (667.5)                & $[$8495.1,    & 10556.7$]$    & 74.8 & 72.8 & 7.1  \\
     & N-t2d200.01      & 57098.6                 & (4452.8)   & 33465.2   & (2001.0)               & $[$20284.5,   & 26982.3$]$    & 56.3 & 73.3 & 11.2 \\
     & N-t2d200.02      & 54895.6                 & (4589.7)   & 35901.8   & (2077.5)               & $[$15539.1,   & 22448.5$]$    & 55.6 & 73.5 & 11.2 \\
     & N-p50-01         & 21431.2                 & (299.5)    & 21027.6   & (711.8)                & $[$-128.1,    & 935.3$]$      & 219.9& 72.3 & 0.9  \\
     & N-p50-02         & 22021.1                 & (275.6)    & 21613.6   & (596.4)                & $[$-42.5,     & 857.5$]$      & 217.3& 72.3 & 0.9  \\
     & N-atp111         & 693.1                   & (10.3)     & 594.0     & (15.4)                 & $[$86.7,      & 111.5$]$      & 98.7 & 72.5 & 4.1  \\
     & N-atp134         & 880.5                   & (18.6)     & 756.3     & (21.3)                 & $[$105.4,     & 143.0$]$      & 83.0 & 72.8 & 5.7  \\
     & N-be75eec\_150   & 1666193.1               & (72310.8)  & 1415208.5 & (58593.5)              & $[$188960.6,  & 313008.6$]$   & 74.0 & 72.9 & 6.9  \\
     & N-be75eec\_250   & 5214413.0               & (166701.4) & 4431271.0 & (109756.7)             & $[$649040.8,  & 917243.2$]$   & 44.9 & 73.7 & 15.5 \\
     & N-be75np\_150    & 3742444.0               & (88500.2)  & 3204878.3 & (80704.1)              & $[$457944.2,  & 617187.2$]$   & 74.2 & 72.9 & 6.9  \\
     & N-be75np\_250    & 10887427.8              & (159642.5) & 9544754.9 & (260523.9)             & $[$1136635.8, & 1548710.0$]$  & 45.3 & 74.0 & 15.5 \\\midrule
   PFSP     & rec05        & 1317.9                  & (25.1)     & 1330.7    & (17.9)                 & $[$-33.4,     & 7.8$]$        & 400.0& 38.9 & 0.1  \\
       & rec13        & 2128.2                  & (25.0)     & 2158.7    & (29.5)                 & $[$-56.2,     & -4.8$]$       & 400.0& 38.9 & 0.1  \\
       & rec19        & 2398.6                  & (19.8)     & 2405.1    & (12.6)                 & $[$-22.3,     & 9.3$]$        & 400.0& 82.1 & 0.3  \\\midrule
    QAP     & kra32        & 115907.0                & (4663.4)   & 116320.0  & (4673.5)               & $[$-4799.3,   & 3973.3$]$     & 400.0& 95.8 & 0.3  \\
        & nug12        & 624.2                   & (26.3)     & 660.4     & (17.3)                 & $[$-57.3,     & -15.1$]$      & 400.0& 14.8 & 0.0  \\
        & nug30        & 7486.8                  & (40.4)     & 7474.0    & (83.9)                 & $[$-50.8,     & 76.4$]$       & 400.0& 84.3 & 0.3  \\
        & tho30        & 192416.0                & (2584.2)   & 191482.0  & (4419.7)               & $[$-2527.0,   & 4395.0$]$     & 400.0& 85.1 & 0.3  \\
